---
title: "Bio302 Home Exam 2024 Part A"
format: 
  html:
    self-contained: true
    code-tools: 
      source: true
---

## Part A - reproducibility

The paper I chose is  "From individual to population level: Temperature and snow cover modulate fledging success through breeding phenology in greylag geese (_Anser anser_)" (Frigerio, Didone et al. (2021). From individual to population level: Temperature and snow cover modulate fledging success through breeding phenology in Greylag geese (Anser anser) [Dataset]. Dryad. https://doi.org/10.5061/dryad.np5hqbztd)

There are several factors evaluated (like productivity of the geese) but the main result I will try to reproduce is the relationship of the first laid egg with the predictors winter temperature and snow depth as well as the relationship between the length of the egg-laying window and the number of breeding pairs. These are presented in figure 3 and table 1 in the original paper. 

```{r}
#| label: read data
library(tidyverse)

geese_data <- readxl::read_xlsx(here::here("Frigerio_Sumasgutner_et_al_Sci_Rep_Supporting_Data.xlsx"), sheet = 2) |> janitor::clean_names()
```

```{r}
#| label: model first egg vs. temperature
mod_temp <- lm(first_egg ~ temp_winter, data = geese_data)
performance::check_model(mod_temp)

broom::augment(mod_temp, interval = "confidence", level = 0.95) |> 
  ggplot(aes(x = temp_winter, y = first_egg)) +
  geom_point() +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = .3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Winter temperature [°C]", y = "1st egg laid") +
  theme_bw()
```

```{r}
#| label: model first egg vs. snow depth
mod_snow <- lm(first_egg ~ snow_average, data = geese_data)
performance::check_model(mod_snow)

broom::augment(mod_snow, interval = "confidence", level = 0.95) |> 
  ggplot(aes(x = snow_average, y = first_egg)) +
  geom_point() +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = .3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Annual snow depth [cm]", y = "1st egg laid") +
  theme_bw()
```

```{r}
#| label: model egg laying time window vs. number of breeding pairs
mod_window <- lm(period ~ pairs, data = geese_data)
performance::check_model(mod_window)

broom::augment(mod_window, interval = "confidence", level = 0.95) |> 
  ggplot(aes(x = pairs, y = period)) +
  geom_point() +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = .3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Nb. breeding pairs", y = "Length egg laying time window") +
  theme_bw()
```


Archive the code and text on GitHub and include the link to your repo in your answer here.

-   Make at least two commits to git.
-   Cite the original paper and data.
-   Make your reproduction as reproducible as possible (i.e. it should be possible for me to reproduce your work after cloning your repo and downloading the data).

1.  Discuss how well you managed to reproduce the part of paper you were working on and what challenges you faced.

3.  Discuss what how well the original paper meets best practice for reproducibility, i.e. what could the authors have done to make their analysis more reproducible, and what they did well.

- not easy to find open data in the first place, databases with papers and corresponding datasets could be better and more papers should have openly accessible datasets
